---
title: "Model 1: A probabilistic model of segment borrowability"
author: "Steven Moran and Elad Eisen"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
bibliography: '../bibliography.bib'  
---

<!-- author: "Steven Moran and Elad Eisen" -->
  
# Overview

This report includes supplementary materials for:

>>> Operationalizing borrowability: A case study from phonological segments

And it uses the R programming language [@R] and the following R libraries [@tidyverse;@knitr].

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
```

# Data

First, load the data from the [CLDF format](https://cldf.clld.org) [@Forkel_etal2018] and combine the tables into single data frames including [PHOIBLE](https://phoible.org) [@MoranMcCloy2019] and [SegBo](https://github.com/segbo-db/segbo) [@segbo].

```{r, message=FALSE}
values <- read_csv('../data/segbo/cldf/values.csv')
languages <- read_csv('../data/segbo/cldf/languages.csv')
segbo <- left_join(values, languages, by=c("Language_ID"="ID"))

values <- read_csv('../data/phoible/cldf/values.csv')
languages <- read_csv('../data/phoible/cldf/languages.csv')
phoible <- left_join(values, languages, by=c("Language_ID"="ID"))
```

How many languages in SEGBO are also in PHOIBLE? There are currently 498 languages in SegBo, of which 199 are *not* in PHOIBLE.

```{r}
tmp1 <- segbo %>% select(Language_ID) %>% distinct()
tmp2 <- phoible %>% select(Language_ID) %>% distinct()
nrow(tmp1)
nrow(tmp1[which(!(tmp1$Language_ID %in% tmp2$Language_ID)),])
```


# Analysis

A segment's borrowability factor `bS` is calculated by @Eisen2019[p.60] as:

$$
b_s = \frac{P_s}{f_s - f_s^2}
$$

where `Ps` is the probability of borrowing a segment and `fs` is its typological frequency. Empirical estimations for the probability of borrowing and the typological frequency are estimated from SegBo and PHOIBLE, respectively.

Get the typological frequency of segments in phoible.

```{r}
num_phoible_inventories <- phoible %>% select(Inventory_ID) %>% distinct() %>% nrow()
phoible_segment_counts <- as.data.frame(table(phoible$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / num_phoible_inventories
```

Get segbo segment frequencies.

```{r}
num_segbo_inventories <- segbo %>% select(Inventory_ID) %>% distinct() %>% nrow()
segbo_segment_counts <- as.data.frame(table(segbo$Value))
segbo_segment_counts <- segbo_segment_counts %>% rename(Segment = Var1, SegboFreq = Freq)
segbo_segment_counts$Segment <- as.character(segbo_segment_counts$Segment)
segbo_segment_counts$SegboCrossFreq <- segbo_segment_counts$SegboFreq / num_segbo_inventories
```

Which segments are in segbo that are not in phoible?

```{r}
segbo_segment_counts[which(!(segbo_segment_counts$Segment %in% phoible_segment_counts$Segment)),]
# 154	halh1238		Mongolian	svantesson2005mongolian ʊai	mand1415 - from Mandarian, perhaps uai
# 285	assa1263	ɹ̤	sans1269 -from Sanskrit
# 304	chra1242	ɨə̯	viet1252-- from Vietnamese
```

Combine SEGBO and PHOIBLE

```{r}
bs_df_new <- left_join(segbo_segment_counts, phoible_segment_counts)
```

Calculate all borrowability scores.

```{r}
bs_df_new$BorrowabilityScore <- bs_df_new$SegboCrossFreq / (bs_df_new$PhoibleCrossFreq - bs_df_new$PhoibleCrossFreq ^ 2)
# write_csv(bs_df_new, 'all_borrowability_scores.csv')
```

Recall that these borrowability scores are not normalized (i.e. 0--1) and that the typological frequencies and borrowing frequencies use two different sets of languages, PHOIBLE and SEGBO samples respectively. That is, some of the languages from SEGBO are not in PHOIBLE, so borrowing frequencies can be in excess of their respective typological frequenies -- causing some rare segments borrowability coefficients to be quite high. For example:

```{r}
bs_df_new %>% arrange(desc(BorrowabilityScore)) %>% head(n=20) %>% select(Segment, BorrowabilityScore, PhoibleCrossFreq, SegboCrossFreq) %>% kable()
```

Here we display the top 20 with a cross-linguistic frequency greater than .05 to get a clearer picture of those segments in both databases. Below we take the intersection of the two, as per Model 2.

```{r}
t <- bs_df_new %>% filter(PhoibleCrossFreq > .05) %>% arrange(desc(BorrowabilityScore)) %>% head(n=20) %>% select(Segment, BorrowabilityScore, PhoibleCrossFreq, SegboCrossFreq)

t <- t %>% rename(Phoneme = Segment, Borrowability = BorrowabilityScore, `Typological Frequency` = PhoibleCrossFreq, `Borrowing Frequency` = SegboCrossFreq)

library(xtable)
# print(xtable(t), include.rownames=FALSE)
t %>% kable()
```


## Random sampling

One problem, however, is that cross-linguistic frequency is treated as *all* the data points and not just a unique list of languages. So for example, segments in either data source (PHOIBLE or SEGBO) are counted multiple times -- consider the four inventories for Akan in phoible.

```{r}
phoible %>% filter(Glottocode == "akan1250") %>% group_by(Inventory_ID, Glottocode, ISO639P3code, Name) %>% summarize(SegmentCount = n()) %>% kable()
```

So, segments that are shared across these inventories are counted more than one for the "same" language.

Let's try random sampling one inventory per language from PHOIBLE and then calculate the borrowability scores. Later we can compare the results between various samples.

```{r}
phoible %>%
    distinct(Inventory_ID, ISO639P3code) %>%
    group_by(ISO639P3code) %>%
    sample_n(1) %>%
    pull(Inventory_ID) ->
    inventory_ids_sampled_one_per_isocode

phoible %>%
    filter(Inventory_ID %in% inventory_ids_sampled_one_per_isocode) ->
    my_sample

num_phoible_inventories <- my_sample %>% select(Inventory_ID) %>% distinct() %>% nrow()
phoible_segment_counts <- as.data.frame(table(my_sample$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / num_phoible_inventories
```

We'll use the same SEGBO counts from above, even though they produce the same problem, i.e. there are multiple reports ("inventories") for the same language.

Combine SEGBO and random PHOIBLE

```{r}
bs_df_random <- left_join(segbo_segment_counts, phoible_segment_counts)
```

Calculate all borrowability scores.

```{r}
bs_df_random$BorrowabilityScore <- bs_df_random$SegboCrossFreq / (bs_df_random$PhoibleCrossFreq - bs_df_random$PhoibleCrossFreq ^ 2)
```

Let's have a look.

```{r}
bs_df_random %>% filter(PhoibleCrossFreq > .05) %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


## Collapse all inventories

In contrast to the approach above, [Model 2](../neighbor-graph_model/A language-graph-based model of segment borrowing.ipynb) simply collapses all of the duplicate inventories into "single" languages, instead of randomizing or bootstrapping the samples. Let's do the same here for PHOIBLE and SEGBO. Then we can compare the original, randomly sampled, and all-inventories-as-one language approaches and the resulting borrowability scores.

```{r}
collapsed_inventories <- phoible %>% group_by(ISO639P3code) %>% select(ISO639P3code, Value)

num_phoible_inventories <- collapsed_inventories %>% select(ISO639P3code) %>% distinct() %>% nrow()
phoible_segment_counts <- as.data.frame(table(collapsed_inventories$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / num_phoible_inventories
```

And do the same for SEGBO.

```{r}
segbo_collapsed <- segbo %>% group_by(ISO639P3code, Value)
num_segbo_inventories <- segbo %>% select(ISO639P3code) %>% distinct() %>% nrow()
segbo_segment_counts <- as.data.frame(table(segbo_collapsed$Value))
segbo_segment_counts <- segbo_segment_counts %>% rename(Segment = Var1, SegboFreq = Freq)
segbo_segment_counts$Segment <- as.character(segbo_segment_counts$Segment)
segbo_segment_counts$SegboCrossFreq <- segbo_segment_counts$SegboFreq / num_segbo_inventories
```

Combine them and get the borrowability score.

```{r}
bs_df_collapsed <- left_join(segbo_segment_counts, phoible_segment_counts)
bs_df_collapsed$BorrowabilityScore <- bs_df_collapsed$SegboCrossFreq / (bs_df_collapsed$PhoibleCrossFreq - bs_df_collapsed$PhoibleCrossFreq ^ 2)
```

Let's have a look.

```{r}
bs_df_collapsed %>% filter(PhoibleCrossFreq > .05) %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


## Intersection of segments in PHOIBLE and SEGBO

As per Model 2's assumptions, here we take the intersection of languages in SEGBO and PHOIBLE. Recall above that we calculate how many languages are in both databases. There are 498 languages in SEGBO and 2177 in PHOIBLE and 199 languages in SEGBO not in PHOIBLE, leaving 299 languages that are shared by both databases.

```{r}
tmp1 <- segbo %>% select(Language_ID) %>% distinct()
tmp2 <- phoible %>% select(Language_ID) %>% distinct()
nrow(tmp1)
nrow(tmp2)
nrow(tmp1[which(!(tmp1$Language_ID %in% tmp2$Language_ID)),])
```

Subset PHOIBLE to just the languages in Segbo.

```{r}
phoible_intersection <- phoible %>% filter(phoible$Language_ID %in% tmp1$Language_ID)
nrow(phoible_intersection %>% distinct(Language_ID))
```

We note again that PHOIBLE contains multiple doculects for the same language.

```{r}
phoible_intersection %>% group_by(Inventory_ID, Glottocode) %>% select(Language_ID, Glottocode) %>% distinct() %>% arrange(Glottocode) %>% head() %>% kable()
```

We simply collapse these different doculects into single phonological inventories. 

```{r}
collapsed_inventories <- phoible_intersection %>% group_by(ISO639P3code) %>% select(ISO639P3code, Value)
```

Nevertheless, we still want the typological frequency of segments from all of PHOIBLE for our borrowability scores.

```{r}
num_phoible_inventories <- phoible %>% select(Inventory_ID) %>% distinct() %>% nrow()
phoible_segment_counts <- as.data.frame(table(phoible$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / num_phoible_inventories
```

We also want just the languages in SEGBO that are in PHOIBLE.

```{r}
segbo_intersection <- segbo %>% filter(Language_ID %in% phoible$Language_ID)
segbo %>% select(Language_ID) %>% distinct() %>% nrow()
segbo_intersection %>% select(Language_ID) %>% distinct() %>% nrow() # Should also be 299 like above
```

We also collapse doculects in SEGBO.

```{r}
segbo_collapsed <- segbo_intersection %>% group_by(ISO639P3code, Value)
```

Now get their segment counts.

```{r}
num_segbo_inventories <- segbo_collapsed %>% select(ISO639P3code) %>% distinct() %>% nrow()
segbo_segment_counts <- as.data.frame(table(segbo_collapsed$Value))
segbo_segment_counts <- segbo_segment_counts %>% rename(Segment = Var1, SegboFreq = Freq)
segbo_segment_counts$Segment <- as.character(segbo_segment_counts$Segment)
segbo_segment_counts$SegboCrossFreq <- segbo_segment_counts$SegboFreq / num_segbo_inventories
```

Combine them.

```{r}
bs_df_intersected <- inner_join(segbo_segment_counts, phoible_segment_counts)
```

Calculate all borrowability scores.

```{r}
bs_df_intersected$BorrowabilityScore <- bs_df_intersected$SegboCrossFreq / (bs_df_intersected$PhoibleCrossFreq - bs_df_intersected$PhoibleCrossFreq ^ 2)
```

Let’s have a look.

```{r}
bs_df_intersected %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


## Accounting for negative data in Model 1

So it turns out that the problem of negative data was present in Model 1 all along: 

1. Either borrowing_relative_frequency should also be computed with num_phoible_inventories in the denominator, or 
2. We should indeed take the intersection of SEGBO and PHOIBLE for typological_relative_frequency as well.

Approach 1:

```{r}
# Let's start with computing borrowing relative frequency with the number of inventories in phoible in the denominator of our borrowability equation.
# bs_df_new$BorrowabilityScore <- bs_df_new$SegboCrossFreq / (bs_df_new$PhoibleCrossFreq - bs_df_new$PhoibleCrossFreq ^ 2)

```

Approach 2:

Let's take the intersection of SEGBO and PHOIBLE for borrowing and typological relative frequency.

```{r}
tmp1 <- segbo %>% select(Language_ID) %>% distinct()
tmp2 <- phoible %>% select(Language_ID) %>% distinct()
nrow(tmp1)
nrow(tmp2)
nrow(tmp1[which(!(tmp1$Language_ID %in% tmp2$Language_ID)),])

shared_glottocodes <- intersect(tmp1$Language_ID, tmp2$Language_ID)
length(shared_glottocodes)
```

Subset PHOIBLE to just the languages in Segbo.

```{r}
phoible_intersection <- phoible %>% filter(phoible$Language_ID %in% shared_glottocodes)
nrow(phoible_intersection %>% distinct(Language_ID))
```

We note again that PHOIBLE contains multiple doculects for the same language.

```{r}
phoible_intersection %>% group_by(Inventory_ID, Glottocode) %>% select(Language_ID, Glottocode) %>% distinct() %>% arrange(Glottocode) %>% head() %>% kable()
```

We simply collapse these different doculects into single phonological inventories. The result is that if two phonological descriptions of the same dialect or language (as determined by their Glottocode) are different, e.g., one author posits 5 vowels and another 10 vowels, i.e., they present evidence for phonemic vowel length, then the resulting collapsed inventories is the union of the two vowel inventories. 

```{r}
collapsed_inventories <- phoible_intersection %>% select(Glottocode, Value) %>% group_by(Glottocode, Value) %>% distinct() %>% ungroup()
nrow(collapsed_inventories %>% select(Glottocode) %>% distinct())
```

We also want just the languages in SEGBO that are in PHOIBLE.

```{r}
segbo_intersection <- segbo %>% filter(Language_ID %in% shared_glottocodes)
segbo_intersection %>% select(Language_ID) %>% distinct() %>% nrow() # Should also be 299 like above
```

We also collapse doculects in SEGBO.

```{r}
segbo_collapsed <- segbo_intersection %>% select(Language_ID, Value) %>% group_by(Language_ID, Value) %>% distinct() %>% ungroup()
segbo_collapsed <- segbo_collapsed %>% rename(Glottocode = Language_ID)
nrow(segbo_collapsed %>% select(Glottocode) %>% distinct())
```

Now we are dealing with two potentially different samples. For example, if segment X occurs in a description (a doculect) in SEGBO, i.e., it is reported as a borrowed segment -- but that doculect is not in PHOIBLE, but another description is that e.g., does not describe borrowed segments -- then our "language" samples between databases differs. The result after applying the conditional probability borrowing equation will result in results outside the range of 0-1. Examples are given in the next section.

To alleivate this issue, we make further assumption. In the few cases that different doculects represent the same dialect of language at the intersection of the two databases (PHOIBLE and SEGBO), borrowed segments are intersected (and if they aren't reported in PHOIBLE added to the inventory) in PHOIBLE. 


```{r}
collapsed_inventories <- rbind(collapsed_inventories, segbo_collapsed)
collapsed_inventories <- collapsed_inventories %>% group_by(Glottocode, Value) %>% distinct() %>% ungroup()
nrow(collapsed_inventories %>% select(Glottocode) %>% distinct())
```

Total inventories.

```{r}
length(unique(segbo_collapsed$Glottocode))
length(unique(collapsed_inventories$Glottocode))

# collapsed_inventories[which(!(unique(collapsed_inventories$ISO639P3code) %in% unique(segbo_collapsed$ISO639P3code))),]
```

Now get their respective segment counts and use the size of the intersected language sample in our borrowability equation.

```{r}
phoible_segment_counts <- as.data.frame(table(collapsed_inventories$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / length(shared_glottocodes)

any(phoible_segment_counts$PhoibleCrossFreq > 1)
any(phoible_segment_counts$PhoibleCrossFreq < 0)
```


```{r}
segbo_segment_counts <- as.data.frame(table(segbo_collapsed$Value))
segbo_segment_counts <- segbo_segment_counts %>% rename(Segment = Var1, SegboFreq = Freq)
segbo_segment_counts$Segment <- as.character(segbo_segment_counts$Segment)
segbo_segment_counts$SegboCrossFreq <- segbo_segment_counts$SegboFreq / length(shared_glottocodes)

any(segbo_segment_counts$PhoibleCrossFreq > 1)
any(segbo_segment_counts$PhoibleCrossFreq < 0)
```

Combine them.

```{r}
bs_df_intersected_merged <- inner_join(segbo_segment_counts, phoible_segment_counts)
```

Calculate borrowability scores.

```{r}
bs_df_intersected_merged$BorrowabilityScore <- bs_df_intersected_merged$SegboCrossFreq / (bs_df_intersected_merged$PhoibleCrossFreq - bs_df_intersected_merged$PhoibleCrossFreq ^ 2)
```

Let's have a look. But we still get values over one...

```{r}
bs_df_intersected_merged %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


Try to repair the algorithm wit

P(contact) = 6 * fs * (1 - fs)

and because of normalization, so:

P(borrowing) = P(contact) * P(borrowing|contact) = 6 * fs * (1 - fs) * bs

and 

bs = Ps / (fs - fs^2) / 6

So, re-calculate borrowability scores.

```{r}
bs_df_intersected_merged$BorrowabilityScore <- bs_df_intersected_merged$SegboCrossFreq / (bs_df_intersected_merged$PhoibleCrossFreq - bs_df_intersected_merged$PhoibleCrossFreq ^ 2) / 6
```

But x / (x-x^2) still tends to infinity as x tends to 1, and surpasses 6 somewhere after x = 0.8 [here]( (https://www.wolframalpha.com/input?i=plot+x+%2F+%28x+-+x%5E2%29+from+0+to+1).

We note that given the nature of the data sets, if we see 1 occurrence of borrowed segment X in SEGBO and also just 1 occurrence of it in PHOIBLE, it means that the only occurrence of that documented sound is borrowed -- which theoretically could be the case  intersected between these two databases. 

And this may actually happen even with the best possible data: if a language borrowed a rare segment from another language and then the latter than quickly lost it, it may be recorded as recent borrowing in one language and not recorded at all in another or anywhere else.

Again, this is marginal, so we can be upfront about this. Or we can apply Laplace smoothing and bump up typological frequencies of all segments by 1 (unless they appear in all languages) to break the ties.

Have a look.

```{r}
bs_df_intersected_merged %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


## Accounting for negative data in Model 1 with typological frequencies from PHOIBLE

Now let's try the same intersection model, but with typological frequencies from ALL of PHOIBLE.

```{r}
num_phoible_inventories <- phoible %>% select(Inventory_ID) %>% distinct() %>% nrow()
phoible_segment_counts <- as.data.frame(table(phoible$Value))
phoible_segment_counts <- phoible_segment_counts %>% rename(Segment = Var1, PhoibleFreq = Freq)
phoible_segment_counts$Segment <- as.character(phoible_segment_counts$Segment)
phoible_segment_counts$PhoibleCrossFreq <- phoible_segment_counts$PhoibleFreq / num_phoible_inventories

any(phoible_segment_counts$PhoibleCrossFreq > 1)
any(phoible_segment_counts$PhoibleCrossFreq < 0)
```

Combine them.

```{r}
bs_df_intersected_merged_all_phoible <- inner_join(segbo_segment_counts, phoible_segment_counts)
```

Calculate borrowability scores with the new formula.

```{r}
bs_df_intersected_merged_all_phoible$BorrowabilityScore <- bs_df_intersected_merged_all_phoible$SegboCrossFreq / (bs_df_intersected_merged_all_phoible$PhoibleCrossFreq - bs_df_intersected_merged_all_phoible$PhoibleCrossFreq ^ 2) / 6
```

Let's have a look. But we still get values over one...

```{r}
bs_df_intersected_merged_all_phoible %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```


# Compare the results

Compare the results (but not the intersection results).

```{r}
bs_df_new_cut <- bs_df_new %>% select(Segment, PhoibleCrossFreq, BorrowabilityScore)
bs_df_collapsed_cut <- bs_df_collapsed %>% select(Segment, BorrowabilityScore)%>% rename(BorrowabilityScore_collapsed = BorrowabilityScore)
bs_df_random_cut <- bs_df_random %>% select(Segment, BorrowabilityScore) %>% rename(BorrowabilityScore_random = BorrowabilityScore)
bs_df_intersected <- bs_df_intersected %>% select(Segment, BorrowabilityScore) %>% rename(BorrowabilityScore_intersection = BorrowabilityScore)
bs_df_intersected_merged <- bs_df_intersected_merged %>% select(Segment, BorrowabilityScore) %>% rename(BorrowabilityScore_no_neg_data = BorrowabilityScore)
bs_df_intersected_merged_all_phoible <- bs_df_intersected_merged_all_phoible %>% select(Segment, BorrowabilityScore) %>% rename(BorrowabilityScore_no_neg_data_all_phoible = BorrowabilityScore)

results <- left_join(bs_df_intersected_merged_all_phoible, bs_df_intersected_merged)
results <- left_join(results, bs_df_new_cut)
results <- left_join(results, bs_df_collapsed_cut)
results <- left_join(results, bs_df_random_cut)
results <- left_join(results, bs_df_intersected)
```

Have a look at all.

```{r}
results %>% arrange(desc(BorrowabilityScore_no_neg_data_all_phoible)) %>% head(n=25) %>% kable()
```

Have a look at typologically frequent segments reported in PHOIBLE.

```{r}
results %>% filter(PhoibleCrossFreq > .05) %>% arrange(desc(BorrowabilityScore)) %>% head(n=25) %>% kable()
```

Let's write the data to disk.

```{r}
write_csv(results, 'borrowability_scores.csv')
```

Compare the segment borrowability rankings.

```{r}
bs_df_new_cut <- bs_df_new %>% filter(PhoibleCrossFreq > .05) %>% select(Segment, PhoibleCrossFreq, BorrowabilityScore) %>% arrange(desc(BorrowabilityScore))

bs_df_new_cut$OriginalRank <- seq.int(nrow(bs_df_new_cut))

bs_df_collapsed_cut <- bs_df_collapsed %>% filter(PhoibleCrossFreq > .05) %>% select(Segment, BorrowabilityScore) %>% arrange(desc(BorrowabilityScore))

bs_df_collapsed_cut$CollapsedRank <- seq.int(nrow(bs_df_collapsed_cut))

bs_df_random_cut <- bs_df_random %>% filter(PhoibleCrossFreq > .05) %>% select(Segment, BorrowabilityScore) %>% arrange(desc(BorrowabilityScore))

bs_df_random_cut$RandomizedRank <- seq.int(nrow(bs_df_random_cut))

results <- left_join(bs_df_new_cut, bs_df_collapsed_cut, by = c("Segment"="Segment"))
results <- left_join(results, bs_df_random_cut, by = c("Segment"="Segment"))
```

Let's have a look.

```{r}
tmp <- results %>% select(-BorrowabilityScore, -BorrowabilityScore.x, -BorrowabilityScore.y)
tmp %>% kable()
```

Let's visualize the differences in ranks that are different. First we need to get the data into a format for plotting.

```{r}
diff_ranks <- tmp %>% select(Segment, OriginalRank, CollapsedRank) %>% filter(OriginalRank != CollapsedRank)

x <- diff_ranks %>% select(Segment, OriginalRank)
x$RankType <- 'OriginalRank'
x <- x %>% rename(Rank = OriginalRank)

y <- diff_ranks %>% select(Segment, CollapsedRank)
y$RankType <- 'CollapsedRank'
y <- y %>% rename(Rank = CollapsedRank)

z <- rbind(x, y)
```

Now we plot using a slope graph.

```{r}
ggplot(data = z, aes(x = RankType, y = Rank, group = Segment)) +
  geom_line(aes(color = Segment, alpha = 1), size = 2) +
  geom_point(aes(color = Segment, alpha = 1), size = 4) +
#  scale_y_discrete(limits = rev(levels(Rank))) +
  #  Labelling as desired
  labs(
    title = "Original versus collapsed inventories",
    # subtitle = "(Mainstreet Research)",
    # caption = "https://www.mainstreetresearch.ca/gap-between-ndp-and-pcs-narrows-while-liberals-hold-steady/"
  )
```

This is quite a bit. Let's plot the deltas of more than 3 rank changes.

```{r}
diff_ranks$delta <- abs(diff_ranks$OriginalRank - diff_ranks$CollapsedRank)
top_deltas <- diff_ranks %>% arrange(desc(delta)) %>% filter(delta > 3)
to_plot <- z %>% filter(Segment %in% top_deltas$Segment)

ggplot(data = to_plot, aes(x = RankType, y = Rank, group = Segment)) +
  geom_line(aes(color = Segment, alpha = 1), size = 2) +
  geom_point(aes(color = Segment, alpha = 1), size = 4) +
#  scale_y_discrete(limits = rev(levels(Rank))) +
  #  Labelling as desired
  labs(
    title = "Original versus collapsed inventories",
    # subtitle = "(Mainstreet Research)",
    # caption = "https://www.mainstreetresearch.ca/gap-between-ndp-and-pcs-narrows-while-liberals-hold-steady/"
  )
```

Also perhaps not so helpful. How about in terms of the more frequent segments with a delta.

```{r}
tmp$OCdelta <- abs(tmp$OriginalRank - tmp$CollapsedRank)
# tmp_by_freq_rank <- tmp %>% filter(OCdelta > 0) %>% arrange(desc(PhoibleCrossFreq, OriginalRank)) %>% head(n=10)
tmp_by_freq_rank <- tmp %>% filter(OCdelta > 0) %>% arrange(desc(PhoibleCrossFreq)) %>% head(n=10)

x <- tmp_by_freq_rank %>% select(Segment, OriginalRank)
x$RankType <- 'OriginalRank'
x <- x %>% rename(Rank = OriginalRank)

y <- tmp_by_freq_rank %>% select(Segment, CollapsedRank)
y$RankType <- 'CollapsedRank'
y <- y %>% rename(Rank = CollapsedRank)

z <- rbind(x, y)

ggplot(data = z, aes(x = RankType, y = Rank, group = Segment)) +
  geom_line(aes(color = Segment, alpha = 1), size = 2) +
  geom_point(aes(color = Segment, alpha = 1), size = 4) +
#  scale_y_discrete(limits = rev(levels(Rank))) +
  #  Labelling as desired
  labs(
    title = "Original versus collapsed inventories",
    # subtitle = "(Mainstreet Research)",
    # caption = "https://www.mainstreetresearch.ca/gap-between-ndp-and-pcs-narrows-while-liberals-hold-steady/"
  )

```


# Figures

Figures for the paper.

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(dplyr)
```

Load the data.

```{r}
segbo_data <- read_csv("model_1_borrowability.csv", col_types = cols(Segment = col_character()))
```


```{r, freq_scatter}
p <- ggplot(segbo_data, aes(x=PHOIBLE_frequency_relative, y=SEGBO_frequency_relative)) +
  # ggtitle(paste('Borrowing frequency vs. typological frequency'))  +
  xlab("Typological frequency") + 
  ylab("Frequency of borrowing")+
  geom_text(label=segbo_data$Segment)+ xlim(0, 1)+ ylim(0, 0.05)

# ggsave(file="freq_scatter.svg", plot=p, width=10, height=6)
# pdf("freq_scatter.pdf", width=10, height=6)
# p
# dev.off()
p
```

Plot.

```{r, freq_scatter_w_lines}
for (i in 1:nrow(segbo_data)) {
  if (segbo_data[i,"PHOIBLE_frequency_relative"]>0.8) {
    segbo_data[i,"group"] <- "5"
  } else if (segbo_data[i,"PHOIBLE_frequency_relative"]>0.6){
    segbo_data[i,"group"] <- "4"
  } else if (segbo_data[i,"PHOIBLE_frequency_relative"]>0.4){
    segbo_data[i,"group"] <- "3"
  } else if (segbo_data[i,"PHOIBLE_frequency_relative"]>0.2){
    segbo_data[i,"group"] <- "2"
  } else {
    segbo_data[i,"group"] <- "1"
  }
}

p <- ggplot(segbo_data, aes(x=PHOIBLE_frequency_relative, y=SEGBO_frequency_relative, color = group)) + #geom_point() + 
  # ggtitle('Borrowing frequency vs. typological frequency') + 
  xlab("Typological frequency") + 
  ylab("Frequency of borrowing")+ 
  geom_text(label=segbo_data$Segment)+ xlim(0, 1)+ ylim(0, 0.05) +
  geom_smooth(method = "lm", formula = y ~ 1, linewidth = 2, se = 0) +
  theme(legend.position = "none") +
  scale_color_manual(values=c("#E69F00", "#009E73", "#D55E00", "#0072B2", "#CC79A7"))

#ggsave(file="freq_scatter_w_lines.svg", plot=p, width=10, height=6)
#pdf("freq_scatter_w_lines.pdf", width=10, height=6)
#p
#dev.off()
p
```

# Other stats

Some stats for the paper.

Load the data.

```{r}
df <- read_csv("model_1_borrowability_w_features.csv", col_types = cols(Segment = col_character()))
```

Get the average of consonants.

```{r}
# c <- df %>% filter(Type == "consonant")
# mean(c$Borrowability)

df %>% filter(Type == "consonant") %>% summarize(mean(Borrowability))
```

Get the average of vowels

```{r}
df %>% filter(Type == "vowel") %>% summarize(mean(Borrowability))
```

Features.

```{r}
table(df$Subtype)
```

Borrowability averages by feature.

```{r}
df %>% filter(Subtype == "affricate") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "approximant") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "fricative") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "lateral") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "nasal") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "plain") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "rhotic") %>% summarize(mean(Borrowability))
df %>% filter(Subtype == "stop") %>% summarize(mean(Borrowability))
```

Borrowability by voiced / voiceless.


```{r}
c_voiceless <- df %>% filter(Type == "consonant") %>% filter(ConsonantVoicing == 0)
mean(c_voiceless$Borrowability)
```

```{r}
c_voiceless <- df %>% filter(Type == "consonant") %>% filter(ConsonantVoicing == 1)
mean(c_voiceless$Borrowability)
```

Wilcoxon rank-sum test for whether consonants or vowels are statistically more significant when comparing mean scores.

```{r}
wilcox.test(Borrowability ~ Type, data = df, conf.int = T)
```

# References
